---
title: "Most damaging events in terms of human and economic damage."
author: "luiyusen97"
date: "7/22/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Abstract

This short analysis finds out the overall most damaging events in terms of population and economic damage, in order to best focus preventive measures.

## Data Processing

Data was downloaded from the links in the url vector. The first URL is the one containing the raw data. No initial transformations were made.

```{r data_processing}
# loading packages
library(tidyverse)

# file downloading DATA PROCESSING
url <- c("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2",
         "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf",
         "https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf")
if (!file.exists("rawdata.csv.bz2")){
    download.file(url = url[1],
                  destfile = "rawdata.csv.bz2")
}
if (!file.exists("rawdata_doc.pdf")){
    download.file(url = url[2],
                  destfile = "rawdata_doc.doc")
}
if (!file.exists("rawdata_faq.pdf")){
    download.file(url = url[3],
                  destfile = "rawdata_faq.pdf")
}

#read data
stormdata <- read.csv("rawdata.csv.bz2")
```

## Population damage

Population damage was taken from the FATALITIES and INJURIES columns of the raw data. Sums were taken of both damage types of each event type, as in figure 2. Negligible event types are those that have overall caused very low population damage in terms of fatalities and injuries. The cut-off point of 1000 was enough to narrow the list of most damaging events down to 14. Figure 1 below shows the 14 most damaging event types.

```{r pop_dmg}
# harmful wrt population hp FATALITIES/INJURIES

# Calculate the sums of fatalities and injuries for each event type
stormdata_orderedevtype <- group_by(stormdata, EVTYPE, .drop = FALSE)
sum_fatalities <- summarise(stormdata_orderedevtype, sum_fat = sum(FATALITIES))
sum_injuries <- summarise(stormdata_orderedevtype, sum_inj = sum(INJURIES))
sum_popdmg <- merge(sum_fatalities, sum_injuries, by = "EVTYPE")

# Remove data entries with negligible values to narrow down selection of event 
# to the most damaging
empty_indices <- vector(mode = "numeric")
for (i in 1:nrow(sum_popdmg)){
    if ((sum_popdmg[i, 2] < 1000) & (sum_popdmg[i, 3] < 1000)){
        empty_indices <- c(empty_indices, i)
    }
}
sum_popdmg <- sum_popdmg[-empty_indices, ]
sum_popdmg <- mutate(sum_popdmg, eventtype = 1:nrow(sum_popdmg))

# find the maximum values for fatalities and injuries, and their corresponding
# event type
max_fat <- sum_popdmg[which(sum_popdmg$sum_fat==max(sum_popdmg$sum_fat)), ]
max_inj <- sum_popdmg[which(sum_popdmg$sum_inj==max(sum_popdmg$sum_inj)), ]

# plot the most damaging events in different colours, and circle the largest 
# values for fatalities and injuries
sum_popdmg_plot <- ggplot(sum_popdmg) +
    geom_point(mapping = aes(x = eventtype, y = sum_fat, colour = "red")) + 
    geom_point(mapping = aes(x = eventtype, y = sum_inj, colour = "blue")) +
    geom_point(shape = 1, mapping = aes(x = max_fat[ , 4], y = max_fat[ , 2])) +
    geom_point(shape = 1, mapping = aes(x = max_inj[ , 4], y = max_inj[ , 3])) +
    labs(title = "Figure 1: Fatalities and Injuries by event type",
         x = "Event type", y = "Fatalities/Injuries") +
    scale_color_manual(labels = c("Fatalities", "Injuries"),
                       values = c("red", "cadetblue3"))
print(sum_popdmg_plot)
```


### Figure 2
```{r figure2}
print(sum_popdmg)
```

This is the processed sum data for population damage.

## Economic damage
The data measured economic damage in terms of the value of property and crops damaged and lost. To find the overall most economically damaging event types, I summed up the damage done over the years for each event type, as in Figure 4. 10000 was the cut-off point to ignore negligible damage values and reduce the number of relevant event types to 31. Figure 3 plots out the damage values and the greatest values are circled in black.

```{r econ_dmg}
# economic damage
# calculate the sums of property and economic damage for each event type
sum_propdmg <- summarise(stormdata_orderedevtype, sum_propdmg = sum(PROPDMG))
sum_cropdmg <- summarise(stormdata_orderedevtype, sum_cropdmg = sum(CROPDMG))
sum_econdmg <- merge(sum_propdmg, sum_cropdmg, by = "EVTYPE")

# remove values that are negligible. Keep the most damaging events
empty_indices <- vector(mode = "numeric")
for (i in 1:nrow(sum_econdmg)){
    if ((sum_econdmg[i, 2] < 10000) & (sum_econdmg[i, 3] < 10000)){
        empty_indices <- c(empty_indices, i)
    }
}
sum_econdmg <- sum_econdmg[-empty_indices, ]
sum_econdmg <- mutate(sum_econdmg, eventtype = 1:nrow(sum_econdmg))

# find the greatest damage values for both property and crop damage
max_propdmg <- sum_econdmg[which(sum_econdmg$sum_propdmg==max(sum_econdmg$sum_propdmg)), ]
max_cropdmg <- sum_econdmg[which(sum_econdmg$sum_cropdmg==max(sum_econdmg$sum_cropdmg)), ]

# plot the most damaging events with circles outlining the greatest values for both types
# of damage
sum_econdmg_plot <- ggplot(sum_econdmg) +
    geom_point(mapping = aes(x = eventtype, y = sum_propdmg, colour = "red")) + 
    geom_point(mapping = aes(x = eventtype, y = sum_cropdmg, colour = "blue")) +
    geom_point(shape = 1, mapping = aes(x = max_propdmg[ , 4], y = max_propdmg[ , 2])) +
    geom_point(shape = 1, mapping = aes(x = max_cropdmg[ , 4], y = max_cropdmg[ , 3])) +
    labs(title = "Figure 3: Property and crop damage by event type",
         x = "Event type", y = "Property/Crop damage") +
    scale_color_manual(labels = c("Property damage", "Crop damage"),
                       values = c("red", "cadetblue3"))
print(sum_econdmg_plot)
```


### Figure 4
```{r figure4}
print(sum_econdmg)
```

This is the processed sum data for economic damage.

## Results

The event type with the greatest overall fatalities and injuries over the years is the `r max_fat[ , 1, drop = TRUE]` with overall fatality numbers of `r max_fat[ , 2, drop = TRUE]` and `r max_fat[ , 3, drop = TRUE]` injuries.
The greatest property damage of `r max_propdmg[ , 2, drop = TRUE]` is caused by the event type of `r max_propdmg[ , 1, drop = TRUE]`, while the greatest crop damage of `r max_cropdmg[ , 3, drop = TRUE]` is caused by the event type `r max_cropdmg[ , 1, drop = TRUE]`.